MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators

### 2024-05-05 rebuttal.txt

First of all, we would like to thank the reviewers for their consideration of our paper and detailed remarks. We have addressed them in our paper revision. The individual answers are in the following. Minor remarks that were simply fixed in the paper and required no additional comments are omitted.

Reviewer #1:

Q: The Boolean signaling modeling paragraph needs to be extended in more details: To me it is not clear, why e.g., the transition rates should be in [0, inf) and not [0,1], since we are talking about probabilities. The concept of asynchronous updating should be made a bit clearer as well. Furthermore, it is often referred to the evaluation of the Boolean formulas as the most expensive computational part. It would help to get an idea, how such a formula would look like, and how they are implemented, to give the reader context about this problem. Furthermore, the word "node" is used quite often in the paper, so it should be made clear, what this refers to exactly.
A: We extended the background section with all your comments. Regarding the node, [0, inf) transition rates and asynchronous updates, we added explanations to the first paragraphs of sections 2.1 and 2.2. In order to add an example of Boolean formula, we explained it in more detail by defining its binary function, which determines the transition rate using a simple rule in equation 5. Having the binary function, we presented its simple example in equation 4. The way how the boolean formulas are implemented in the code is described in the first paragraph of section 3.1, simply referring to it as a recursive traversal of an expression tree. We believe that for the sake of giving the context to boolean logic evaluation, we included enough information. If the reader wants to dive into details, we extended the readme in maboss.GPU repository with a detailed description of formulas implementation.

@@

Q: It is stated that synthetic models with up to 1000 nodes were created. The synthetic models themselves are available, however, the description of the chosen models is also important to be mentioned in the paper: I would suggest a table that lists the most important features of the different models (size, nodes, formula sizes, how many different models were actually taken into account, ..) that were created (this table can also include the characteristics of the 3 real-world models to put in perspective). 
A: We included the proposed table in Results section and extended the first paragraph in the Benchmarking Methodology to describe the table fields.

Q: The legend in the figures needs to be more specific to be able to interpret the figures better: e.g. how many synthetic models? (are the 19 points the respective models?)
A: Yes, the x-axis shows multiple synthetic models of various sizes. We added this to Figure 2.

Q: In Figure1, the High-End NVIDIA TESLA is either slower or as fast as the consumer grade laptop GPU. Is this purely due to the compilation time (as Figure 2 would suggest, that it makes a difference)? 
A: You are right. You can see for yourself by comparing gpu_out_real.csv and gpu_out_real_a100.csv files in the artifact. We are unsure why the compilation takes longer for high-end GPU than for laptop GPU, after all, it is a proprietary piece of software. Perhaps the loading of shared libraries takes a different amount of time for the operating systems... We added a note in the Performance of MaBoSS.GPU section.

Q: Please ellaborate more on the last point in Figure 5, that seems to be an outlier. Which synthetic model is this? Is it also found in the earlier figures? 
A: Firstly, we included additional information about the synthetic model in the figure caption. Secondly, we analyzed the datapoint, and it is the same case as in the previous MPI figure - the problem is distributed into so many small tasks that the actual cost of MPI communication becomes the bottleneck. We are sorry that we did not include this information in the first version - initially, the figure was plotted in a different way, which did not show this drastic speedup difference; after replotting, the datapoint went unnoticed.

### 2024-05-06 Article.tex
Added changes for Q1 and Q3

Biological systems are large and complex, and understanding their internal behavior remains critical for designing new therapies for complex diseases such as cancer.
A crucial approach in this endeavor is building computational models from existing knowledge and analyzing them to find intervention points and to predict the efficacy of new treatments.
Many different frameworks have been used to describe biological systems, from quantitative systems of differential equations to more qualitative approaches such as Boolean models.
While the former seems more adapted to represent complex behavior, such as non-linear dependencies, the latter is being increasingly used because of its capability to analyze very large systems.
Many Boolean models have been built to describe biological systems to tackle a variety of problems: from understanding fundamental properties of cell cycle~\cite{faure2006cellcycle,sizek2019boolean} to advanced properties of cancer~\cite{fumia_carcinogenesis_2013,montagud2022prostate}.

Historically, the task of building Boolean models involved reading an extensive amount of literature and summarizing it in a list of essential components and their interactions.
More recently, database listings of such interactions~\cite{licata2020signor,turei2016omnipath} and experimental information retrieval techniques on a bigger number of components were subjected to many advancements. This resulted in the design of automatic methods for Boolean formulae inference from the constraints encoded in the knowledge and the experimental data~\cite{10.1093/bioinformatics/btaa484,chevalier2020synthesis,benevs2023boolean}, allowing construction of large Boolean models.
While this effort faces many challenges, we believe it is a promising way to study the large-scale complexity of biological systems.
However, in order to analyze the dynamic properties of such large Boolean models, we need to develop efficiently scalable simulation tools.

Here, we present adaptations of MaBoSS~\cite{stoll2012continuous, stoll2017maboss} --- a stochastic Boolean simulator that performs estimations of state probability trajectories based on Markov chains -- to modern HPC computing architectures, which provide significant speedups of the computation, thus allowing scrutinization and analysis of much larger boolean models.
The main contributions comprise two new implementations of MaBoSS:
\begin{itemize}
    \item MaBoSS.GPU, a GPU-accelerated implementation of MaBoSS, which is designed to exploit the computational power of massively parallel GPU hardware.
    \item MaBoSS.MPI, a parallel implementation of MaBoSS which can scale to multinode environments, such as large CPU clusters.
\end{itemize}

@@

MaBoSS algorithm simulates the above process to produce stochastic \emph{trajectories}: sequences of states $S^0, S^1, \dots, S^k$ and time points $t^0 < t^1 < \dots < t^k$ where $t^0 = 0$ and $S^0$ is the initial state, and for each $i \in \{0, \dots, k-1\}$, $S^i$ transitions to $S^{i+1}$ at time $t^{i+1}$. The simulation ends either by a timeout when reaching the maximal allowed time, or by reaching a \emph{fixed point} state with no outgoing transitions. The algorithm for a single iteration of the trajectory simulation is given explicitly in Algorithm~\ref{alg:iter}, which is the direct application of the Gillespie stochastic simulation algorithm~\cite{gillespie1976general} (or kinetic Monte-Carlo~\cite{young1966monte}) on the Boolean state space.

@@

As mentioned at the beginning of the section, if the trajectory does not reach a fixed point, the simulation is stopped after the maximal allowed time. This is a common scenario, especially when some trajectories form circles, i.e., when a model has \emph{circular steady states}. A circular steady state is usually not directly visible from the state probability line charts; Stoll et al.~\cite{stoll2012continuous} proposed methods to detect them (such as plotting the state and transition entropies), but we do not discuss the methods further in this paper for the sake of brevity.

>>>>

Biological systems are large and complex, and understanding their internal behavior remains critical for designing new therapies for complex diseases such as cancer.
A crucial approach in this endeavor is building computational models from existing knowledge and analyzing them to find intervention points and to predict the efficacy of new treatments~\cite{brodland2015computational,bongrand2023understanding}.
Many different frameworks have been used to describe biological systems, from quantitative systems of differential equations to more qualitative approaches such as Boolean models~\cite{machado2011modeling}.
While the former seems more adapted to represent complex behavior, such as non-linear dependencies, the latter is being increasingly used because of its capability to analyze very large systems.
Many Boolean models have been built to describe biological systems to tackle a variety of problems: from understanding fundamental properties of cell cycle~\cite{faure2006cellcycle,sizek2019boolean} to advanced properties of cancer~\cite{fumia_carcinogenesis_2013,wooten_sclg,montagud2022prostate}.

Historically, the task of building Boolean models involved reading an extensive amount of literature and summarizing it in a list of essential components and their interactions.
More recently, database listings of such interactions~\cite{licata2020signor,turei2016omnipath} and experimental information retrieval techniques on a bigger number of components were subjected to many advancements. Combined with the design of automatic methods for Boolean formulae inference from the constraints encoded in the knowledge and the experimental data~\cite{10.1093/bioinformatics/btaa484,chevalier2020synthesis,benevs2023boolean,prugger_unsupervised}, these new developments allows the construction of large Boolean models.
While this effort faces many challenges, we believe it is a promising way to study the large-scale complexity of biological systems.
However, in order to analyze the dynamic properties of such large Boolean models, we need to develop efficiently scalable simulation tools.

Here, we present adaptations of MaBoSS~\cite{stoll2012continuous, stoll2017maboss} --- a stochastic Boolean simulator that performs estimations of state probability trajectories based on Gillespie stochastic simulation algorithm~\cite{gillespie1976general} -- to modern HPC computing architectures, which provide significant speedups of the computation, thus allowing scrutinization and analysis of much larger boolean models. In particular, the problem of properly quantifying low abundant phenotypes~\cite{gillespie2013perspective} can now be tackled by making more realistic the large number of simulation needed to cover the space of possible trajectories.
The main contributions comprise two new implementations of MaBoSS:
\begin{itemize}
    \item MaBoSS.GPU, a GPU-accelerated implementation of MaBoSS, which is designed to exploit the computational power of massively parallel GPU hardware.
    \item MaBoSS.MPI, a parallel implementation of MaBoSS which can scale to multinode environments, such as large CPU clusters.
\end{itemize}

@@

MaBoSS algorithm simulates the above process to produce stochastic \emph{trajectories}: sequences of states $S^0, S^1, \dots, S^k$ and time points $t^0 < t^1 < \dots < t^k$ where $t^0 = 0$ and $S^0$ is the initial state, and for each $i \in \{0, \dots, k-1\}$, $S^i$ transitions to $S^{i+1}$ at time $t^{i+1}$. The simulation ends either by a timeout when reaching the maximal allowed time, or by reaching a \emph{fixed point} state with no outgoing transitions. The algorithm for a single iteration of the trajectory simulation is given explicitly in Algorithm~\ref{alg:iter}, which is the direct application of the Gillespie stochastic simulation algorithm on the Boolean state space.

@@

As mentioned at the beginning of the section, if the trajectory does not reach a fixed point, the simulation is stopped after the maximal allowed time. This is a common scenario, especially when some trajectories form cycles, i.e., when a model has \emph{cyclic attractor}, also known as limit cycles. A limit cycle is usually not directly visible from the state probability line charts; Stoll et al.~\cite{stoll2012continuous} proposed methods to detect them (such as plotting the state and transition entropies), but we do not discuss the methods further in this paper for the sake of brevity.
